---
title: "Sex Differences in Immune Response to COVID-19"
author: "Mallory Ellingson" 
Date: "June 24th, 2020"
output: html_notebook
---

# Data import & data cleaning

## Loading necessary libaries

```{r}
library(corrplot)
library(dplyr)
```

## Master Spreadsheet (06/24/20)

```{r}
#Importing data (06/24/20)
master <- readxl::read_xlsx("/Users/malloryellingson/Documents/COVID/ORIGINAL_FINAL_MasterFlow_datasheet_200625_PHI_removed.xlsx")
```

```{r}

# 06/24/20

#Fixing study IDS 
master$hcw.or.pt <- substr(master$ID, 1, 3)
master$pt <- 1
master$pt[master$hcw.or.pt == "HCW"] <- 0
master$newID[master$pt == 1] <- substr(master$ID[master$pt == 1], 1, 5)
master$newID[master$pt == 0] <- substr(master$ID[master$pt == 0], 1, 6)
covid <- master

```


### Data Cleaning

```{r}


#Recoding gender as binary numeric variable
covid$sex[covid$gender == "M"] <- 0
covid$sex[covid$gender == "F"] <- 1

#Creating indicator for ICU status
covid$icu.stat <- 0
covid$icu.stat[covid$ICU == 2] <- 1

#Getting age information for all patient
covid$age[covid$age == "≥ 90"] <- "90"
covid$age <- as.numeric(as.character(covid$age))

#Categorical BMI
covid$bmi.cat[covid$BMI < 18.5] <- 1
covid$bmi.cat[covid$BMI >= 18.5 & covid$BMI <= 24.9] <- 2
covid$bmi.cat[covid$BMI > 24.9 & covid$BMI <= 29.9] <- 3
covid$bmi.cat[covid$BMI > 29.9 & covid$BMI <= 35.0] <- 4
covid$bmi.cat[covid$BMI > 35.0] <- 5

#Creating a time dependent variable for tocilizumab treatment
covid$DayfromTociiStart <- as.character(covid$DayfromTociiStart)
covid$DayfromTociiStart[covid$DayfromTociiStart == "NoInfo"] <- 99
covid$DayfromTociiStart <- as.numeric(covid$DayfromTociiStart)

covid$on.tocii[covid$DayfromTociiStart <= 0] <- 0
covid$on.tocii[covid$DayfromTociiStart > 0 & covid$DayfromTociiStart <= 14] <- 1
covid$on.tocii[covid$DayfromTociiStart > 14 & covid$DayfromTociiStart != 99] <- 0
covid$on.tocii[is.na(covid$DayfromTociiStart)] <- 0
covid$on.tocii[covid$DayfromTociiStart == 99] <- NA

#Creating a time dependent variable for corticosteroid treatment
covid$DaysfromCSStart <- as.character(covid$DaysfromCSStart)
covid$DaysfromCSEnd <- as.character(covid$DaysfromCSEnd)

covid$DaysfromCSStart[covid$DaysfromCSStart == "NoInfo"] <- 99
covid$DaysfromCSEnd[covid$DaysfromCSEnd == "NoInfo"] <- 99

covid$DaysfromCSEnd <- as.numeric(covid$DaysfromCSEnd)
covid$DaysfromCSStart <- as.numeric(covid$DaysfromCSStart)

covid$on.cs <- 0
covid$on.cs[covid$DaysfromCSEnd <= 0 & covid$DaysfromCSStart > 0] <- 1
covid$on.cs[covid$DaysfromCSEnd == 99] <- NA

#Setting tocii & CS and ICU treatment to 0 for HCW
covid$on.tocii[covid$pt == 0] <- 0
covid$on.cs[covid$pt == 0] <- 0
covid$in.icu[covid$pt == 0] <- 0

```

### Creating subsetted datasets

```{r}
#Dataset limited just to first observations
covid.base <- covid[covid$obs == 1,]

#Dataset limited to just baseline observations but patients
covid.pt.all.base <- covid.base[covid.base$pt == 1,]

#Dataset limited to first obs, non-ICU patients, no tocii and no steroids (HCW & patients)
covid.nonicu <- covid.sex[covid.sex$icu.stat == 0 & covid.sex$tocii == 0 & covid.sex$cs == 0,]

#Same dataset as above, but just patients
covid.pt.base <- covid.nonicu[covid.nonicu$pt == 1,]

#Only PT
covid.pt <- covid[covid$pt == 1,]
covid.pt <- covid.pt[covid.pt$Exclusion != 1,]

#Patients at each timepoint

covid1 <- covid.pt[covid.pt$TP == 1,]
covid2 <- covid.pt[covid.pt$TP == 2,]
covid3 <- covid.pt[covid.pt$TP == 3,]
covid4 <- covid.pt[covid.pt$TP == 4,]
covid5 <- covid.pt[covid.pt$TP == 5,]
covid6 <- covid.pt[covid.pt$TP == 6,]

#Calculating mean and SD of BMI for all groups

covid1.f <- covid1[covid1$sex == 1,]
#mean(covid1.f$BMI)
#sd(covid1.f$BMI)

covid1.m <- covid1[covid1$sex == 0,]
#mean(covid1.m$BMI)
#sd(covid1.m$BMI)

covid1.f.a <- covid1.f[covid1.f$CohortA == 1,]
#mean(covid1.f.a$BMI)
#sd(covid1.f.a$BMI)

covid1.m.a <- covid1.m[covid1.m$CohortA == 1,]
#mean(covid1.m.a$BMI)
#sd(covid1.m.a$BMI)

covid1.hcw <- covid[covid$hcw.or.pt == "HCW",]
covid1.hcw <- covid1.hcw[covid1.hcw$TP == 1,]

covid1.hcw.f <- covid1.hcw[covid1.hcw$sex == 1,]

```

### Export for SAS

```{r}
#See SAS code for Data Analysis
write.csv(covid, "/Users/malloryellingson/Documents/COVID/covidforsas.csv")
```


## Patient Trajectory Data - Data Cleaning

(Patient trajectory data is identical to data above, just limited to Cohort A and first time point observations with a column for the patient trajectory. Details on how patient trajectory was calculated can be found in the manuscript)

### Importing trajectory data

```{r}
cohortA.traj <- readxl::read_xlsx("/Users/malloryellingson/Documents/COVID/200624_Cohort_A_1st_time_point_MKE.xlsx")
```

### Data cleaning and merging

```{r}
#Limiting to relevant variables
covid.traj <- cohortA.traj[, 1:440]

#Making sex numerical; 

covid.traj$sex[covid.traj$gender == "F"] <- 1
covid.traj$sex[covid.traj$gender == "M"] <- 0

```


```{r}
write.csv(covid.traj, "/Users/malloryellingson/Documents/COVID/trajforsas.csv")
```

## Exploring the data

```{r}
#Type I or Type III Interferon Response
hist(covid.base$IFNa2)
hist(covid.base$IFNL2orIL28A)
hist(covid.base$IFNy)

#T-Cell Phenotypes
hist(covid.pt.all.base$CD4Naivenum)
hist(covid.pt.all.base$CD4Temnum)
hist(covid.pt.all.base$CD4Tcmnum)
hist(covid.pt.all.base$NaiveCD8Tnum)
hist(covid.pt.all.base$CD8Temranum)
hist(covid.pt.all.base$CD8Tcmnum)
hist(covid.pt.all.base$CD8TofCD3)
hist(covid.pt.all.base$CD4TCRactofCD3)
hist(covid.pt.all.base$CD8TCRactofCD8)
hist(covid.pt.all.base$CD8TexofCD3)
hist(covid.pt.all.base$CD8IFNyofCD3)
hist(covid.pt.all.base$CD8GzBofCD3)

#Viral Load
hist(covid.pt.all.base$Nptiter)

#Antibody Response
hist(covid.pt.all.base$AntiS1IgGA450)
hist(covid.pt.all.base$AntiS1IgMA450)

#Inflammatory Cytokines
hist(covid.pt.all.base$FLT3L)
hist(covid.pt.all.base$CCL2orMCP1)
hist(covid.pt.all.base$CCL8OrMCP2)
hist(covid.pt.all.base$CCL5orRANTES)
hist(covid.pt.all.base$MCSF)
hist(covid.pt.all.base$CCL3orMIP1a)
hist(covid.pt.all.base$IL15)
hist(covid.pt.all.base$TRAIL)
hist(covid.pt.all.base$CXCL13OrBCA1)

#Eosinophils and IgE
hist(covid.pt.all.base$Eosnum)
hist(covid.pt.all.base$IgE)

#Additional ELISA results
hist(covid.pt.all.base$IL8orCXCL8)
hist(covid.pt.all.base$IL18)
hist(covid.pt.all.base$sCD40L)

#Monocytes
hist(covid.pt.all.base$ncMonoofLive)
hist(covid.pt.all.base$intMonoofLive)
hist(covid.pt.all.base$cMonoofLive)

```

### Association between outcome variables (immune response measurements) and covariates of interest

Just patients (too many missing values for HCW)

```{r}
#Limiting dataset to a set of variables of interest
covid.corr <- covid.pt.base[,c("newID", "age", "sex", "DaysfromSxOnset", "ethnicity", "bmi.cat", "riskfac", "IFNa2", "IFNL2orIL28A", "CD4Naivenum", "CD4Temnum", "CD4Tcmnum", "NaiveCD8Tnum", "CD8Temofnum", "CD8Tcmnum", "AntiS1IgGA450", "AntiS1IgMA450", "Nptiter", "FLT3L", "CCL2orMCP1", "CCL8OrMCP2", "CCL5orRANTES", "MCSF", "CCL3orMIP1a", "IL15", "TRAIL", "CXCL13OrBCA1", "Eosnum", "IgE")]

#Converting any variables read as 'factors' to numeric
covid.corr$CCL2orMCP1 <- as.numeric(as.character(covid.corr$CCL2orMCP1))
covid.corr$CCL5orRANTES <- as.numeric(as.character(covid.corr$CCL5orRANTES))
covid.corr$IFNL2orIL28A <- as.numeric(as.character(covid.corr$IFNL2orIL28A))

```


```{r}
#Age and different immune response measurements (patients)

#Type I or Type III interferons
cor.test(covid.corr$IFNa2, covid.corr$age) 
cor.test(covid.corr$IFNL2orIL28A, covid.corr$age) 

#T Cell Phenotypes
cor.test(covid.corr$CD4Naivenum, covid.corr$age)
cor.test(covid.corr$CD4Temnum, covid.corr$age)
cor.test(covid.corr$CD4Tcmnum, covid.corr$age)
cor.test(covid.corr$NaiveCD8Tnum, covid.corr$age)
cor.test(covid.corr$CD8Temofnum, covid.corr$age)
cor.test(covid.corr$CD8Tcmnum, covid.corr$age)

#Antibodies
cor.test(covid.corr$AntiS1IgGA450, covid.corr$age)
cor.test(covid.corr$AntiS1IgMA450, covid.corr$age)

#Viral load
cor.test(covid.corr$Nptiter, covid.corr$age)

#Inflammatory cytokine response
cor.test(covid.corr$FLT3L, covid.corr$age)
cor.test(covid.corr$CCL2orMCP1, covid.corr$age)
cor.test(covid.corr$CCL8OrMCP2, covid.corr$age)
cor.test(covid.corr$CCL5orRANTES, covid.corr$age)
cor.test(covid.corr$MCSF, covid.corr$age)
cor.test(covid.corr$CCL3orMIP1a, covid.corr$age)
cor.test(covid.corr$IL15, covid.corr$age)
cor.test(covid.corr$TRAIL, covid.corr$age)
cor.test(covid.corr$CXCL13OrBCA1, covid.corr$age)

#Eosinophils and IgE
cor.test(covid.corr$Eosnum, covid.corr$age)
cor.test(covid.corr$IgE, covid.corr$age)
```

```{r}
#Time since symptom onset and immune response measurements (patients)

#Type I or Type III interferons
cor.test(covid.corr$IFNa2, covid.corr$DaysfromSxOnset) 
cor.test(covid.corr$IFNL2orIL28A, covid.corr$DaysfromSxOnset)

#T Cell Phenotypes
cor.test(covid.corr$CD4Naivenum, covid.corr$DaysfromSxOnset)
cor.test(covid.corr$CD4Temnum, covid.corr$DaysfromSxOnset)
cor.test(covid.corr$CD4Tcmnum, covid.corr$DaysfromSxOnset)
cor.test(covid.corr$NaiveCD8Tnum, covid.corr$DaysfromSxOnset)
cor.test(covid.corr$CD8Temofnum, covid.corr$DaysfromSxOnset)
cor.test(covid.corr$CD8Tcmnum, covid.corr$DaysfromSxOnset)

#Antibodies
cor.test(covid.corr$AntiS1IgGA450, covid.corr$DaysfromSxOnset)
cor.test(covid.corr$AntiS1IgMA450, covid.corr$DaysfromSxOnset)

#Viral load
cor.test(covid.corr$Nptiter, covid.corr$DaysfromSxOnset)

#Inflammatory cytokine response
cor.test(covid.corr$FLT3L, covid.corr$DaysfromSxOnset)
cor.test(covid.corr$CCL2orMCP1, covid.corr$DaysfromSxOnset)
cor.test(covid.corr$CCL8OrMCP2, covid.corr$DaysfromSxOnset)
cor.test(covid.corr$CCL5orRANTES, covid.corr$DaysfromSxOnset)
cor.test(covid.corr$MCSF, covid.corr$DaysfromSxOnset)
cor.test(covid.corr$CCL3orMIP1a, covid.corr$DaysfromSxOnset)
cor.test(covid.corr$IL15, covid.corr$DaysfromSxOnset)
cor.test(covid.corr$TRAIL, covid.corr$DaysfromSxOnset)
cor.test(covid.corr$CXCL13OrBCA1, covid.corr$DaysfromSxOnset)

#Eosinophils and IgE
cor.test(covid.corr$Eosnum, covid.corr$DaysfromSxOnset)
cor.test(covid.corr$IgE, covid.corr$DaysfromSxOnset)
```

```{r}
#Ethnicity
covid.corr$ethnicity <- as.factor(covid.corr$ethnicity)

#Type I or Type III interferons
res.aov <- aov(IFNa2 ~ ethnicity, data = covid.corr)
summary(res.aov)

res.aov <- aov(IFNL2orIL28A ~ ethnicity, data = covid.corr)
summary(res.aov)

#T Cell Phenotypes
res.aov <- aov(CD4Naivenum ~ ethnicity, data = covid.corr)
summary(res.aov)

res.aov <- aov(CD4Temnum ~ ethnicity, data = covid.corr)
summary(res.aov)

res.aov <- aov(CD4Tcmnum~ ethnicity, data = covid.corr)
summary(res.aov)

res.aov <- aov(NaiveCD8Tnum ~ ethnicity, data = covid.corr)
summary(res.aov)

res.aov <- aov(CD8Temofnum ~ ethnicity, data = covid.corr)
summary(res.aov)

res.aov <- aov(CD8Tcmnum ~ ethnicity, data = covid.corr)
summary(res.aov)

#Antibodies
res.aov <- aov(AntiS1IgGA450 ~ ethnicity, data = covid.corr)
summary(res.aov)

res.aov <- aov(AntiS1IgMA450 ~ ethnicity, data = covid.corr)
summary(res.aov)

#Viral Load
res.aov <- aov(Nptiter ~ ethnicity, data = covid.corr)
summary(res.aov)

#Inflammatory Cytokine Resposne
res.aov <- aov(FLT3L ~ ethnicity, data = covid.corr)
summary(res.aov)

res.aov <- aov(CCL2orMCP1 ~ ethnicity, data = covid.corr)
summary(res.aov)

res.aov <- aov(CCL8OrMCP2 ~ ethnicity, data = covid.corr)
summary(res.aov)

res.aov <- aov(CCL5orRANTES ~ ethnicity, data = covid.corr)
summary(res.aov)

res.aov <- aov(MCSF ~ ethnicity, data = covid.corr)
summary(res.aov)

res.aov <- aov(CCL3orMIP1a ~ ethnicity, data = covid.corr)
summary(res.aov)

res.aov <- aov(IL15 ~ ethnicity, data = covid.corr)
summary(res.aov)

res.aov <- aov(TRAIL ~ ethnicity, data = covid.corr)
summary(res.aov)

res.aov <- aov(CXCL13OrBCA1 ~ ethnicity, data = covid.corr)
summary(res.aov)

#Eosinophils and IgE
res.aov <- aov(Eosnum ~ ethnicity, data = covid.corr)
summary(res.aov)

res.aov <- aov(IgE ~ ethnicity, data = covid.corr)
summary(res.aov)

```

```{r}
#BMI
covid.corr$ethnicity <- as.factor(covid.corr$bmi.cat)

#Type I or Type III interferons
res.aov <- aov(IFNa2 ~ bmi.cat, data = covid.corr)
summary(res.aov)

res.aov <- aov(IFNL2orIL28A ~ bmi.cat, data = covid.corr)
summary(res.aov)

#T Cell Phenotypes
res.aov <- aov(CD4Naivenum ~ bmi.cat, data = covid.corr)
summary(res.aov)

res.aov <- aov(CD4Temnum ~ bmi.cat, data = covid.corr)
summary(res.aov)

res.aov <- aov(CD4Tcmnum~ bmi.cat, data = covid.corr)
summary(res.aov)

res.aov <- aov(NaiveCD8Tnum ~ bmi.cat, data = covid.corr)
summary(res.aov)

res.aov <- aov(CD8Temofnum ~ bmi.cat, data = covid.corr)
summary(res.aov)

res.aov <- aov(CD8Tcmnum ~ bmi.cat, data = covid.corr)
summary(res.aov)

#Antibodies
res.aov <- aov(AntiS1IgGA450 ~ bmi.cat, data = covid.corr)
summary(res.aov)

res.aov <- aov(AntiS1IgMA450 ~ bmi.cat, data = covid.corr)
summary(res.aov)

#Viral Load
res.aov <- aov(Nptiter ~ bmi.cat, data = covid.corr)
summary(res.aov)

#Inflammatory Cytokine Resposne
res.aov <- aov(FLT3L ~ bmi.cat, data = covid.corr)
summary(res.aov)

res.aov <- aov(CCL2orMCP1 ~ bmi.cat, data = covid.corr)
summary(res.aov)

res.aov <- aov(CCL8OrMCP2 ~ bmi.cat, data = covid.corr)
summary(res.aov)

res.aov <- aov(CCL5orRANTES ~ bmi.cat, data = covid.corr)
summary(res.aov)

res.aov <- aov(MCSF ~ bmi.cat, data = covid.corr)
summary(res.aov)

res.aov <- aov(CCL3orMIP1a ~ bmi.cat, data = covid.corr)
summary(res.aov)

res.aov <- aov(IL15 ~ bmi.cat, data = covid.corr)
summary(res.aov)

res.aov <- aov(TRAIL ~ bmi.cat, data = covid.corr)
summary(res.aov)

res.aov <- aov(CXCL13OrBCA1 ~ bmi.cat, data = covid.corr)
summary(res.aov)

#Eosinophils and IgE
res.aov <- aov(Eosnum ~ bmi.cat, data = covid.corr)
summary(res.aov)

res.aov <- aov(IgE ~ bmi.cat, data = covid.corr)
summary(res.aov)
```

```{r}
#Presence of a COVID risk factor

#Type I or Type III interferon respones
t.test(covid.corr$IFNa2, covid.corr$riskfac)
t.test(covid.corr$IFNL2orIL28A, covid.corr$riskfac)

#T Cell Phenotypes
t.test(covid.corr$CD4Naivenum, covid.corr$riskfac)
t.test(covid.corr$CD4Temnum, covid.corr$riskfac)
t.test(covid.corr$CD4Tcmnum, covid.corr$riskfac)
t.test(covid.corr$NaiveCD8Tnum, covid.corr$riskfac)
t.test(covid.corr$CD8Temofnum, covid.corr$riskfac)
t.test(covid.corr$CD8Tcmnum, covid.corr$riskfac)

#Antibodies
t.test(covid.corr$AntiS1IgGA450, covid.corr$riskfac)
t.test(covid.corr$AntiS1IgMA450, covid.corr$riskfac)

#Viral load
t.test(covid.corr$Nptiter, covid.corr$riskfac)

#Inflammatory cytokine response
t.test(covid.corr$FLT3L, covid.corr$riskfac)
t.test(covid.corr$CCL2orMCP1, covid.corr$riskfac)
t.test(covid.corr$CCL8OrMCP2, covid.corr$riskfac)
t.test(covid.corr$CCL5orRANTES, covid.corr$riskfac)
t.test(covid.corr$MCSF, covid.corr$riskfac)
t.test(covid.corr$CCL3orMIP1a, covid.corr$riskfac)
t.test(covid.corr$IL15, covid.corr$riskfac)
t.test(covid.corr$TRAIL, covid.corr$riskfac)
t.test(covid.corr$CXCL13OrBCA1, covid.corr$riskfac)

#Eosinophils and IgE
t.test(covid.corr$Eosnum, covid.corr$riskfac)
t.test(covid.corr$IgE, covid.corr$riskfac)
```



